Finiteness in Cubical Type Theory

### 2020-06-01 sections/finiteness-predicates.tex
cleaner format

  The support list for is \(\left[  \right]\), \(\left[ \text{tt} \right]\), and
  \(\left[ \text{false}, \text{true} \right]\), respectively, and the cover
  proof is a function that returns the index of the supplied element.
\end{proof}
\begin{agdalisting}
  The code in our formalisation of these three proofs is quite simple.
  Just as an example, the proof that \(\mathbf{Bool}\) is split enumerable is as
  follows:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
\end{agdalisting}

With the most basic simple types out of the way, the obvious next choice is the
(non-dependent) sums and products: \(\uplus\) and \(\times\).
Both of these types can be constructed from the \emph{dependent} sum, however,
so that is the type we will prove finite.
From that we can derive a much wider array of finiteness proofs.

@@

  \begin{agdalisting}
    This pattern of applying a function to each element in a list and
    concatenating the result is of course well-known in functional programming,
    and is in fact the pattern that makes lists a monad.
    While this insight isn't strictly relevant to our work here, it does mean
    the implementation of this function can use Agda's do notation, resulting
    in the following extremely clean implementation:
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sup-sigma}
  \end{agdalisting}

  To prove that this support list does in fact cover the entirety of the type
  \(\Sigma \; A \; U\), we note that any element of type \(\Sigma \; A \; U\)
  must have a first component in the support list of \(\mathcal{E}!_A\), and its
  second component must be in the result of applying \(\mathcal{E}!_U\) to that
  first element (since that support list contains every element of type
  \(U(x)\)).
  Therefore, the pair itself must be in our constructed support list.
\end{proof}
\subsection{Derivations}
We have a way to construct finiteness proofs, and a semiring-like toolbox to
combine them.
What we're now interested in is what we can \emph{derive} from them.

@@

Next, we will show that split enumerability is equivalent to the existence of a
split surjection from a finite prefix of the natural numbers:
\begin{lemma} \label{split-enum-is-split-surj}
  \begin{equation}
    \mathcal{E}!(A) \simeq \Sigma (n : \mathbb{N}) , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{E}!(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \in \mathit{xs}
    && \text{def.~\ref{split-enum-def} }(\mathcal{E}!)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{fib}_{\text{snd}(\mathit{xs})}(x)
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{sp-surj-eqn} (sp-surj)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{sp-surj}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \twoheadrightarrow ! \; A )
    && \text{eqn.~\ref{sp-surj-arrow-eqn} } (\twoheadrightarrow!) \; \qedhere
  \end{align*}
\end{proof}
\begin{agdalisting}
In our formalisation, the proof is a single line:
\ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj}
The only step which isn't definitional equality is the reassociation of
\(\Sigma\).
\ExecuteMetaData[agda/Data/Sigma/Properties.tex]{reassoc}
\end{agdalisting}
(The simplicity of this proof, by the way, is why we preferred the
container-based definition of lists over the traditional one.)

@@

To specify that a value must exist uniquely in HoTT we can use the concept of a
\emph{contraction}.
\begin{definition}[Unique Membership] \label{uniq-memb-def}
  \begin{equation}
    x \inunique \mathit{xs} \coloneqq \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{definition}
With this we can define manifest Bishop finiteness:
\begin{definition}[Manifest Bishop Finiteness]  \label{bish-def}
  \begin{equation}
    \mathcal{B}(A) \coloneqq \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \inunique xs
  \end{equation}
\end{definition}
The only difference between manifest Bishop finiteness and split enumerability
is the membership term: here we require unique membership (\(\inunique\)),
rather than simple membership (\(\in\)).

@@

We have now proved that every manifestly Bishop finite type is split enumerable,
and vice versa.
While the types are not \emph{equivalent} (there are more split enumerable
proofs than there are manifest Bishop finite proofs), they are of equal power,
so any closure proof we have on one can be transferred to the other.
In particular, it means that manifest Bishop finiteness is closed under
\(\Sigma\).
\subsection{From Manifest Bishop Finiteness to Equivalence}
We have seen that split enumerability was in fact a split-surjection in
disguise.
We will now see that manifest Bishop finiteness is in fact an \emph{equivalence}
in disguise.
\emph{equivalence} with \(\mathbf{Fin}\).
\begin{lemma} \label{bishop-equiv}
  \begin{equation}
    \mathcal{B}(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{B}(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \inunique \mathit{xs}
    && \text{def.~\ref{bish-def} }(\mathcal{B})
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(x \in \mathit{xs})
    && \text{eqn.~\ref{uniq-memb-def} } (\inunique)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(\text{fib}_{\text{snd}(\mathit{xs})}(x))
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{is-equiv-def} (isEquiv)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{isEquiv}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \simeq A )
    && \text{eqn.~\ref{equiv-def} } (\simeq) \; \qedhere
  \end{align*}
\end{proof}
This proof is almost identical\footnotemark to the proof for
lemma~\ref{split-enum-is-split-surj}: it reveals that
enumeration-based finiteness predicates are simply another perspective on
relation-based ones.
\footnotetext{
  Unfortunately in our formalisation this proof cannot be a single line: for
  performance reasons \(\simeq\) is defined as a record type with eta-equality
  disabled, instead of the definition here which uses \(\Sigma\).
}

@@

To remedy the problem, we will use propositional truncation
(def.~\ref{prop-trunc}).
\begin{definition}[Cardinal Finiteness]
  \begin{equation}
    \mathcal{C}(A) \coloneqq \lVert \mathcal{B}(A) \rVert \simeq \lVert \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \simeq A) \rVert
  \end{equation}
\end{definition}
At first glance, it might seem that we lose any useful properties we could
derive from \(\mathcal{B}\).
Luckily, this is not the case: by eliminator \ref{elim-prop-coh} of
def.~\ref{prop-trunc}, we need only show that the output is uniquely determined.

>>>>

  The support list for is \(\left[  \right]\), \(\left[ \text{tt} \right]\), and
  \(\left[ \text{false}, \text{true} \right]\), respectively, and the cover
  proof is a function that returns the index of the supplied element.
\end{proof}
\begin{marginfigure}%
  \begin{dottedbox}%
    % The code in our formalisation of these three proofs is quite simple.
    % Just as an example, the proof that \(\mathbf{Bool}\) is split enumerable is as
    % follows:
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
  \end{dottedbox}
\end{marginfigure}
With the most basic simple types out of the way, the obvious next choice is the
(non-dependent) sums and products: \(\uplus\) and \(\times\).
Both of these types can be constructed from the \emph{dependent} sum, however,
so that is the type we will prove finite.
From that we can derive a much wider array of finiteness proofs.

@@

  To prove that this support list does in fact cover the entirety of the type
  \(\Sigma \; A \; U\), we note that any element of type \(\Sigma \; A \; U\)
  must have a first component in the support list of \(\mathcal{E}!_A\), and its
  second component must be in the result of applying \(\mathcal{E}!_U\) to that
  first element (since that support list contains every element of type
  \(U(x)\)).
  Therefore, the pair itself must be in our constructed support list.
\end{proof}

\begin{agdalisting}
  This pattern of applying a function to each element in a list and
  concatenating the result is of course well-known in functional programming,
  and is in fact the pattern that makes lists a monad.
  While this insight isn't strictly relevant to our work here, it does mean
  the implementation of this function can use Agda's do notation, resulting
  in the following extremely clean implementation:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sup-sigma}
\end{agdalisting}
\subsection{Derivations}
We have a way to construct finiteness proofs, and a semiring-like toolbox to
combine them.
What we're now interested in is what we can \emph{derive} from them.

@@

We will now see that split enumerability is in fact a split surjection in
another form:
\begin{lemma} \label{split-enum-is-split-surj}%
  A proof of split enumerability is equivalent to a split surjection from a
  finite prefix of the natural numbers.
  \begin{equation}%
    \mathcal{E}!(A) \simeq \Sigma (n : \mathbb{N}) , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{E}!(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \in \mathit{xs}
    && \text{def.~\ref{split-enum-def} }(\mathcal{E}!)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{fib}_{\text{snd}(\mathit{xs})}(x)
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{sp-surj-eqn} (sp-surj)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{sp-surj}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \twoheadrightarrow ! \; A )
    && \text{eqn.~\ref{sp-surj-arrow-eqn} } (\twoheadrightarrow!) \; \qedhere
  \end{align*}
\end{proof}
\begin{agdalisting}
In our formalisation, the proof is a single line:
\ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj}
The only step which isn't definitional equality is the reassociation of
\(\Sigma\).
\ExecuteMetaData[agda/Data/Sigma/Properties.tex]{reassoc}
\end{agdalisting}
(The simplicity of this proof, by the way, is why we preferred the
container-based definition of lists over the traditional one.)

@@

To specify that a value must exist uniquely in HoTT we can use the concept of a
\emph{contraction}.
\begin{definition}[Unique Membership] \label{uniq-memb-def}
  Unique list membership is defined in terms of list membership: it is a
  contraction of it.
  \begin{equation}
    x \inunique \mathit{xs} \coloneqq \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{definition}
With this we can define manifest Bishop finiteness:
\begin{definition}[Manifest Bishop Finiteness]  \label{bish-def}
  A type is manifest Bishop finite if there exists a list which contains each
  value in the type once.
  \begin{equation}
    \mathcal{B}(A) \coloneqq \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \inunique xs
  \end{equation}
\end{definition}
The only difference between manifest Bishop finiteness and split enumerability
is the membership term: here we require unique membership (\(\inunique\)),
rather than simple membership (\(\in\)).

@@

We have now proved that every manifestly Bishop finite type is split enumerable,
and vice versa.
While the types are not \emph{equivalent} (there are more split enumerable
proofs than there are manifest Bishop finite proofs), they are of equal power,
so any closure proof we have on one can be transferred to the other.
In particular, it means that manifest Bishop finiteness is closed under
\(\Sigma\).
\subsection{From Manifest Bishop Finiteness to Equivalence}
We have seen that split enumerability was in fact a split-surjection in
disguise.
We will now see that manifest Bishop finiteness is in fact an \emph{equivalence}
in disguise.
\emph{equivalence} with \(\mathbf{Fin}\).
\begin{lemma} \label{bishop-equiv}
  Manifest bishop finiteness is equivalent to an equivalence to a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{B}(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \inunique \mathit{xs}
    && \text{def.~\ref{bish-def} }(\mathcal{B})
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(x \in \mathit{xs})
    && \text{eqn.~\ref{uniq-memb-def} } (\inunique)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(\text{fib}_{\text{snd}(\mathit{xs})}(x))
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{is-equiv-def} (isEquiv)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{isEquiv}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \simeq A )
    && \text{eqn.~\ref{equiv-def} } (\simeq) \; \qedhere
  \end{align*}
\end{proof}
This proof is almost identical\footnotemark to the proof for
lemma~\ref{split-enum-is-split-surj}: it reveals that
enumeration-based finiteness predicates are simply another perspective on
relation-based ones.
\footnotetext{
  Unfortunately in our formalisation this proof cannot be a single line: for
  performance reasons \(\simeq\) is defined as a record type with eta-equality
  disabled, instead of the definition here which uses \(\Sigma\).
}

@@

To remedy the problem, we will use propositional truncation
(def.~\ref{prop-trunc}).
\begin{definition}[Cardinal Finiteness]
  A type \(A\) is cardinally finite if there exists a propositionally truncated
  proof that \(A\) is manifest Bishop finite or equivalent to a finite prefix of
  the natural numbers.
  \begin{equation}
    \mathcal{C}(A) \coloneqq \lVert \mathcal{B}(A) \rVert \simeq \lVert \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \simeq A) \rVert
  \end{equation}
\end{definition}
At first glance, it might seem that we lose any useful properties we could
derive from \(\mathcal{B}\).
Luckily, this is not the case: by eliminator \ref{elim-prop-coh} of
def.~\ref{prop-trunc}, we need only show that the output is uniquely determined.
